ProSolo: Accurate Variant Calling from Single Cell DNA Sequencing Data

### 2021-09-09 manuscript/supplement.tex
update manuscript, scripts and figures to final submission (last reviewer comment addressed, editorial requests addressed)

\usepackage{hyperref}
\usepackage{tocstyle}
% Define Supplement prefix for numbered references
\newcommand{\supplpref}{S~}
% Set Supplement prefix for numbered references
\renewcommand{\thetable}{\supplpref\arabic{table}}
\renewcommand{\thesection}{\supplpref\arabic{section}}
\renewcommand{\thesubsection}{\supplpref\arabic{section}.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\supplpref\arabic{section}.\arabic{subsection}.\arabic{subsubsection}}
\renewcommand{\thefigure}{\supplpref\arabic{figure}}
\renewcommand{\theequation}{\supplpref\arabic{equation}}

@@

\usepackage{xr-hyper}

@@

\author{David Lähnemann, Johannes Köster, Ute Fischer, \\ Arndt Borkhardt, Alice McHardy, Alexander Schönhuth}

@@

\textbf{Whole genome sequencing of almost identical kin-cells from a cell-line \citep{dong_accurate_2017}.}
The first dataset is from the publication of a previous variant caller for single cell sequencing data generated using MDA, SCcaller (\cite{dong_accurate_2017}; Figure~\ref{fig:datasets}A).
They started with one initial cell from a cell line and grew it in several steps (a complete overview of their setup is given in Figure 1 of \cite{dong_accurate_2017}).
After an initial mini-expansion, they selected one single cell as the founder for their secondary IL expansion into 20-30 cells.
From this they extracted two single cells (IL-11 and IL-12) that were sequenced after multiple displacement amplification (MDA).
Both cells have a minimum coverage of five of at least 80\% of genomic sites (with only about 90\% of the used reference genome represented, as this includes a largely uncovered chromosome Y and alternative haplotype contigs which would only be covered if the particular haplotype is present), with IL-11 having a more even coverage that leads to more sites with a usable coverage (Figure~\ref{fig:coverage-dists}A).
The rest of the kindred cells from that IL clone was then used as a bulk sequencing sample without MDA (IL-1C), showing a coverage profile similar to the whole genome amplified single cells (Figure~\ref{fig:coverage-dists}A).
We generated the ground truth genotypes for this dataset from the kindred clone IL-1C, which should be only very few cell divisions away from the single cells, and thus have almost no difference in the somatic mutations acquired.
The ground truth genotype was generated separately for variant sites and homozygous reference sites:
for variant sites, GATK HaplotypeCaller was run, followed by variant quality score recalibration; for homozygous reference sites, we used bcftools mpileup to identify sites with a coverage above 25 but no alternative nucleotide present.
This bulk sample (IL-1C) was only used as a ground truth and was not provided as input to any of the software compared here.
Three more distant clones (C1, C2, C3 in Figure~\ref{fig:datasets}A), grown from the cells after the initial mini-expansion, were jointly used as a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
With three separate bulk sequencing runs, their joint coverage of about 50 for 80\% of genomic sites is much higher than for the MDA single cells and the unamplified IL-1C bulk (Clones in Figure~\ref{fig:coverage-dists}A).\\

@@

\textbf{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
For the second benchmarking dataset, Blood was taken from a patient after informed consent and granulocytes were sorted out in several steps: Initially, CD3- cells were captured via Magnetic-Activated Cell Sorting (MACS); among them, CD66b+ granulocytes were filtered out via Fluorescence Activated Cell Sorting (FACS) (Figures~\ref{fig:datasets}B and \ref{fig:FACS-scheme}).
Single cells from that cell population were isolated using a Fluidigm C1 microfluidics chip and subjected to multiple displacement amplification (MDA) using the GE Healthcare GenomiPhi Kit in its version adapted to that microfluidic chip.
With a panel of 16 custom primer pairs covering different genes across chromosomes used in a quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
For those cells, we performed whole exome capture using the Agilent SureSelectXT Human All Exon V5+UTRs Kit.
The paired-end sequencing library (2x101bp) was then prepared with the TruSeq V3 Kit and sequenced on a HiSeq 2500.
From the remaining sorted cell population, we extracted bulk DNA and submitted it to the same whole exome capture and DNA sequencing without MDA.\\

@@

For the ground truth of this dataset, we could leverage previously published bulk whole exome sequencing data from the same person, their parents and three siblings (\cite{hoell_constitutional_2014}; Figure~\ref{fig:datasets}B).
For ground truth alternative allele calls, we ran three pedigree-aware variant callers on the family whole exome data.
To be able to run them reproducibly, we made all three of them available via bioconda \citep{gruning_bioconda:_2018}:
\begin{itemize}
  \item BEAGLE 4.0 \citep{browning_improving_2013}, bioconda version: {\ttfamily beagle=4.0\_06Jun17=1}
  \item polymutt \citep{li_likelihood-based_2012}, bioconda version: {\ttfamily polymutt=0.18=0}
  \item FamSeq in MCMC ({\ttfamily -method 3}) mode \citep{peng_rare_2013,peng_famseq:_2014}), bioconda version: {\ttfamily famseq=1.0.3=0}
\end{itemize}
From their calls, we created a consensus by only including sites where either all callers agree on a genotype or where a maximum of one caller has a \emph{missing} (as opposed to different genotype) call, while the other two callers agree on the genotype (Figure~\ref{fig:granulocytes-ground-truth}).
The use of the full pedigree ensures high confidence genotypes---however, please bear in mind that these represent the germline genotype.
Single cells will have accumulated somatic genotypes that differ from the germline and we are bound to mis-interpret those as false positive calls in our ground truth comparison.
At the same time, the number of such somatic variants will be small compared to the number of germline variants, and should hence keep our overall ground truth comparison valid.\\

@@

\textbf{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC) \citep{wang_clonal_2014}.}
The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available.
For each, 16 cells and a bulk sample were whole exome sequenced (Figure~\ref{fig:datasets}C).
For the 32 single cells, library preparation was performed on isolated single nuclei \citep{wang_clonal_2014}.
In addition, \cite{wang_clonal_2014} provide a list of 373 non-synonymous clonal and 94 non-synonymous subclonal somatic variants (Supplementary Tables 6 and 7 in \cite{wang_clonal_2014}) whose presence they validated with deep targeted duplex sequencing.
As they provide a duplex sequencing based alternative allele frequency, we examined the alternative allele frequency distribution of the non-synonymous clonal variants, stratified by the specified zygosity (Figure~\ref{fig:wang-clonals-zygosity}).
This clearly shows two alternative allele frequency peaks, one at around $0.4$ and the other at roughly double the alternative allele frequency around $0.7$.
These two peaks should correspond to heterozygous ($0.4$) and homozygous alternative variants ($0.7$), respectively.
As these are supposed to be clonal variants in the tumor, the remaining fraction of $0.3$ beyond the homozygous alternative variants peak (from $0.7$ to $1$) should be contamination with normal cells (without these somatic variants).
This indicates a duplex sample tumor purity of around $0.7$.
Unexpectedly, we also see the $0.7$ peak in variants that \cite{wang_clonal_2014} classified as heterozygous and the $0.4$ peak in variants that \cite{wang_clonal_2014} classified as homozygous alternative (Figure~\ref{fig:wang-clonals-zygosity}).
It is unclear from \cite{wang_clonal_2014} and its supplementary material how the original variant classifications were generated, we assume that this is based on the genotypes obtained for the single cells.
But as the single cells were genotyped with standard variant calling software (GATK UnifiedGenotyper), they very likely suffer from considerable allelic dropout and we thus reclassified the clonal variants' genotypes based on the duplex sequencing.
If the alternative allele frequency determined by the ultra-deep duplex sequencing was below $0.6$, we classified the variant as heterozygous, and as homozygous alternative if this frequency was above (or equal to) $0.6$ (Figure~\ref{fig:wang-clonals-zygosity}).
Then, depending on the single cell, we used two different ground truths for two separate evaluations:
\begin{enumerate}
    \item For the tumor single cells, we constructed a ground truth from the normal bulk sample (as the tumor bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    Please bear in mind that this ground truth will mostly only contain germline variants and only few somatic mutations that appeared early enough in the patient's life to rise to high somatic frequencies.
    Thus, similar caveats as for the germline ground truth in the granulocyte dataset apply.
    To mitigate this to a certain degree, we added the 373 clonal validated somatic variants, as these are expected to be present in all tumor single cells.
    \item For the normal single cells, we constructed a ground truth from the tumor bulk sample (as the normal bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    To reduce the presence of false-positive somatic variants in this ground truth, we removed all 467 validated somatic variants (clonal and subclonal) from the ground truth.
\end{enumerate}
Further, we determined the of the validated somatic variants for all software run on this dataset.

@@

\begin{figure}[!tpb]
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.9\linewidth}
  \includegraphics[height=42ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot.pdf} \newline
 \end{minipage}
 \caption{
  Global view of the precision-recall plots for alternative allele calling across different software modes and filtering thresholds (for zoomed in view on the most interesting areas of the plot, see main manuscript Figure~\ref{fig:alt-calling_prec-rec}).
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
  \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline
  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequence at the whole exome level. \newline \footnotesize
  Used thresholds are not comparable between tools.
  They were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
  }
 \label{fig:alt-calling_prec-rec_global}
\end{figure}

The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset: all methods achieve high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec_global}A and \ref{fig:alt-calling_prec-rec}A), at recall rates of 45\% and higher. 
In comparison with all other tools, ProSolo achieves the most striking increases in recall of nearly 10\%.
For example, for a precision above .99, its maximum recall is .766 compared to .705 for MonoVar, .687 for SCIPhI and .610 for SCcaller.
SCAN-SNV achieves up to .992 precision, but only .0001 recall.
This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants\footnote{E.g. see this comment in the issues of the respective repository: \url{https://github.com/parklab/scan-snv/issues/5##issuecomment-628107370}}.

Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset.
This dataset is considerably more challenging than the first one.
See Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B for respective results.
For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
On this dataset, SCAN-SNV's recall increased to .0016\% at a decreased maximum precision of .897.
Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
At the sites with somatic mutations in single cells, which SCAN-SNV focuses on, this ground truth will instead contain the germline genotype and will incorrectly classify alternative alleles as false positives.
Due to this effect, we also expect the calculated precision of all the other tools to be an underestimate, but against the backdrop of large numbers of sites where the single cells harbour the germline genotype, the effect will be much smaller.
The same ground truth caveat also applies---inversely---for the recall of SCIPhI's sensitive mode and ProSolo's imputation mode.
Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
While both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation, we expect that a number of imputed sites were incorrectly classified as true positives with this germline ground truth.
As this will lead to an overestimation of the recall, we have excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of this dataset---however, their results are nevertheless displayed in all respective figures.

@@

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_per-cell_prec-rec.pdf}
 \caption{
 Cell-specific precision-recall plots across different filtering threshold levels for the whole exome dataset.\newline \footnotesize
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline
  Used thresholds are not comparable between tools.
  They were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
 \label{fig:per-cell_prec-rec_Laehnemann2017}
\end{figure}

@@

  \item SCIPhI calls nine heterozygous mutations in 12 or 15 cells, while the maximum frequency for any of the heterozygous subclonal mutations detected in the deeply sequenced tumor bulk was 0.233, more in the range of the maximum number of six out of 16 cells that \cite{wang_clonal_2014} found any of the subclonal variants in.
  This suggests that SCIPhI's model over-imputes mutations by aggregating information across cells.

@@

\subsection{Further Software and Resources used}

>>>>

\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage{tocstyle}
% Define Supplement prefix for numbered references
\newcommand{\supplpref}{S~}
% Set Supplement prefix for numbered references
\renewcommand{\thetable}{\supplpref\arabic{table}}
\renewcommand{\thesection}{\supplpref\arabic{section}}
\renewcommand{\thesubsection}{\supplpref\arabic{section}.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\supplpref\arabic{section}.\arabic{subsection}.\arabic{subsubsection}}
\renewcommand{\thefigure}{\supplpref\arabic{figure}}
\renewcommand{\theequation}{\supplpref\arabic{equation}}

@@

@@

\author{David Lähnemann, Johannes Köster, Ute Fischer, \\ Arndt Borkhardt, Alice C. McHardy, Alexander Schönhuth}

@@

\paragraph{Whole genome sequencing of almost identical kin-cells from a cell-line \citep{dong_accurate_2017}.}
The first dataset is from the publication of a previous variant caller for single cell sequencing data generated using MDA, SCcaller (\cite{dong_accurate_2017}; Figure~\ref{fig:datasets}a).
They started with one initial cell from a cell line and grew it in several steps (a complete overview of their setup is given in Figure 1 of \cite{dong_accurate_2017}).
After an initial mini-expansion, they selected one single cell as the founder for their secondary IL expansion into 20-30 cells.
From this they extracted two single cells (IL-11 and IL-12) that were sequenced after multiple displacement amplification (MDA).
Both cells have a minimum coverage of five of at least 80\% of genomic sites (with only about 90\% of the used reference genome represented, as this includes a largely uncovered chromosome Y and alternative haplotype contigs which would only be covered if the particular haplotype is present), with IL-11 having a more even coverage that leads to more sites with a usable coverage (Figure~\ref{fig:coverage-dists}A).
The rest of the kindred cells from that IL clone was then used as a bulk sequencing sample without MDA (IL-1C), showing a coverage profile similar to the whole genome amplified single cells (Figure~\ref{fig:coverage-dists}A).
We generated the ground truth genotypes for this dataset from the kindred clone IL-1C, which should be only very few cell divisions away from the single cells, and thus have almost no difference in the somatic mutations acquired.
The ground truth genotype was generated separately for variant sites and homozygous reference sites:
for variant sites, GATK HaplotypeCaller was run, followed by variant quality score recalibration; for homozygous reference sites, we used bcftools mpileup to identify sites with a coverage above 25 but no alternative nucleotide present.
This bulk sample (IL-1C) was only used as a ground truth and was not provided as input to any of the software compared here.
Three more distant clones (C1, C2, C3 in Figure~\ref{fig:datasets}a), grown from the cells after the initial mini-expansion, were jointly used as a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
With three separate bulk sequencing runs, their joint coverage of about 50 for 80\% of genomic sites is much higher than for the MDA single cells and the unamplified IL-1C bulk (Clones in Figure~\ref{fig:coverage-dists}A).\\

@@

\paragraph{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
For the second benchmarking dataset, Blood was taken from a patient after informed consent and granulocytes were sorted out in several steps: Initially, CD3- cells were captured via Magnetic-Activated Cell Sorting (MACS); among them, CD66b+ granulocytes were filtered out via Fluorescence Activated Cell Sorting (FACS) (Figures~\ref{fig:datasets}b and \ref{fig:FACS-scheme}).
Single cells from that cell population were isolated using a Fluidigm C1 microfluidics chip and subjected to multiple displacement amplification (MDA) using the GE Healthcare GenomiPhi Kit in its version adapted to that microfluidic chip.
With a panel of 16 custom primer pairs covering different genes across chromosomes (Table~\ref{tab:selection-primers}) used in a quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
For those cells, we performed whole exome capture using the Agilent SureSelectXT Human All Exon V5+UTRs Kit.
The paired-end sequencing library (2x101bp) was then prepared with the TruSeq V3 Kit and sequenced on a HiSeq 2500.
From the remaining sorted cell population, we extracted bulk DNA and submitted it to the same whole exome capture and DNA sequencing without MDA.\\

\begin{table}
    \centering
    \caption{
    Primers used for selection of well amplified single cell samples via quantitative real-time PCR.
    These were taken from Table S3 in \cite{zong_genome-wide_2012}.
    }
    \small
    \begin{tabular}{lll}
      \toprule
      chromosome  &  forward primer & reverse primer \\
      \toprule
      chr1  & AGGAAAGGCATACTGGAGGGACAT & TTAGGGATGGCACCACACTCTTGA \\
      \midrule
      chr2  & TCCCAGAGAAGCATCCTCCATGTT & CACCACACTGCCTCAAATGTTGCT \\
      \midrule
      chr3  & TCAAGTTGCCAGCTGTGGCTGTAT & AGAAGGGCATTTCCTGTCAGTGGA \\
      \midrule
      chr4  & ATGGGCAAATCCAGAAGAGTCCAG & CCATTCACTTCCTTGGAAAGGTAGCC \\
      \midrule
      chr5  & AATAGCGTGCAGTTCTGGGTAGCA & TTCACATCCTGGGAGGAACAGCAT \\
      \midrule
      chr6  & TGAATGCCAGGGTGAGACCTTTGA & TGTTCATTATCCCACGCCAGGACT \\
      \midrule
      chr7  & ACCAAAGGAAAGCCAGCCAGTCTA & ACTCCACAGCTCCCAAGCATACAA \\
      \midrule
      chr9  & TCCCAGCTCTCTCTCTTGCATCTT & AGTGAAGCTGGTGTATGCAGAGGT \\
      \midrule
      chr12 & AGAGGGCTGCTTTATGCAGGTG   & CTACATTTGGGTCTTTGCTGCCATG \\
      \midrule
      chr13 & AGCAGCCCCAGGCAGAT        & CGGAGAGGACGGTCACGTTTAC \\
      \midrule
      chr14 & CGGAGAGGACGGTCACGTTTAC   & CGTGGGAGTTTTGAAATGCGATGT \\
      \midrule
      chr15 & CCTGTCTCTGCTCCTGCG       & TGCACACATGCACAGTGGAG \\
      \midrule
      chr16 & CTCCAAGGTTCTGCAGCCTC     & GGTATGACTACACATTCAGGCTGG \\
      \midrule
      chr17 & GTGGTACATAGTGCATGGTCCG   & GGCGACATACCCCAACTTCATAAG \\
      \midrule
      chr18 & CGTTCTTAGGACCAAAGGGCTG   & CCAGCATCCATGTCTCTGCAC \\
      \midrule
      chr19 & GCCCAGAGCGCCTGA          & CCAGCCCCTGGACCACT \\
      \bottomrule
    \end{tabular}
    \label{tab:selection-primers}
\end{table}

@@

For the ground truth of this dataset, we could leverage previously published bulk whole exome sequencing data from the same person, their parents and three siblings (\cite{hoell_constitutional_2014}; Figure~\ref{fig:datasets}b).
For ground truth alternative allele calls, we ran three pedigree-aware variant callers on the family whole exome data.
To be able to run them reproducibly, we made all three of them available via bioconda \citep{gruning_bioconda:_2018}:
\begin{itemize}
  \item BEAGLE 4.0 \citep{browning_improving_2013}, bioconda version: {\ttfamily beagle=4.0\_06Jun17=1}
  \item polymutt \citep{li_likelihood-based_2012}, bioconda version: {\ttfamily polymutt=0.18=0}
  \item FamSeq in MCMC ({\ttfamily -method 3}) mode \citep{peng_rare_2013,peng_famseq:_2014}), bioconda version: {\ttfamily famseq=1.0.3=0}
\end{itemize}
From their calls, we created a consensus by only including sites where either all callers agree on a genotype or where a maximum of one caller has a \emph{missing} (as opposed to different genotype) call, while the other two callers agree on the genotype (Figure~\ref{fig:granulocytes-ground-truth}).
The use of the full pedigree ensures high confidence genotypes---however, please bear in mind that these represent the germline genotype.
Single cells will have accumulated somatic genotypes that differ from the germline and we are bound to mis-interpret those as false positive calls in our ground truth comparison.
At the same time, the number of such somatic variants will be small compared to the number of germline variants, and should hence keep our overall ground truth comparison valid.\\

@@

\paragraph{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC) \citep{wang_clonal_2014}.}
The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available.
For each, 16 cells and a bulk sample were whole exome sequenced (Figure~\ref{fig:datasets}c).
For the 32 single cells, library preparation was performed on isolated single nuclei \citep{wang_clonal_2014}.
In addition, \cite{wang_clonal_2014} provide a list of 373 non-synonymous clonal and 94 non-synonymous subclonal somatic variants (Supplementary Tables 6 and 7 in \cite{wang_clonal_2014}) whose presence they validated with deep targeted duplex sequencing.
As they provide a duplex sequencing based alternative allele frequency, we examined the alternative allele frequency distribution of the non-synonymous clonal variants, stratified by the specified zygosity (Figure~\ref{fig:wang-clonals-zygosity}).
This clearly shows two alternative allele frequency peaks, one at around $0.4$ and the other at roughly double the alternative allele frequency around $0.7$.
These two peaks should correspond to heterozygous ($0.4$) and homozygous alternative variants ($0.7$), respectively.
As these are supposed to be clonal variants in the tumor, the remaining fraction of $0.3$ beyond the homozygous alternative variants peak (from $0.7$ to $1$) should be contamination with normal cells (without these somatic variants).
This indicates a duplex sample tumor purity of around $0.7$.
Unexpectedly, we also see the $0.7$ peak in variants that \cite{wang_clonal_2014} classified as heterozygous and the $0.4$ peak in variants that \cite{wang_clonal_2014} classified as homozygous alternative (Figure~\ref{fig:wang-clonals-zygosity}).
It is unclear from \cite{wang_clonal_2014} and its supplementary material how the original variant classifications were generated, we assume that this is based on the genotypes obtained for the single cells.
But as the single cells were genotyped with standard variant calling software (GATK UnifiedGenotyper), they very likely suffer from considerable allelic dropout and we thus reclassified the clonal variants' genotypes based on the duplex sequencing.
If the alternative allele frequency determined by the ultra-deep duplex sequencing was below $0.6$, we classified the variant as heterozygous, and as homozygous alternative if this frequency was above (or equal to) $0.6$ (Figure~\ref{fig:wang-clonals-zygosity}).
Then, depending on the single cell, we used two different ground truths for two separate evaluations:
\begin{enumerate}
    \item For the tumor single cells, we constructed a ground truth from the normal bulk sample (as the tumor bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    Please bear in mind that this ground truth will mostly only contain germline variants and only few somatic mutations that appeared early enough in the patient's life to rise to high somatic frequencies.
    Thus, similar caveats as for the germline ground truth in the granulocyte dataset apply.
    To mitigate this to a certain degree, we added the 373 clonal validated somatic variants, as these are expected to be present in all tumor single cells.
    \item For the normal single cells, we constructed a ground truth from the tumor bulk sample (as the normal bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    To reduce the presence of false-positive somatic variants in this ground truth, we removed all 467 validated somatic variants (clonal and subclonal) from the ground truth.
\end{enumerate}
Further, we determined the of the validated somatic variants for all software run on this dataset.

@@

\begin{figure}[!tpb]
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.9\linewidth}
  \includegraphics[height=42ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot.pdf} \newline
 \end{minipage}
 \caption{
  Global view of the precision-recall plots for alternative allele calling across different software modes and filtering thresholds (for zoomed in view on the most interesting areas of the plot, see main manuscript Figure~\ref{fig:alt-calling_prec-rec}).
  \textbf{a} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
  \textbf{b} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
  $^\text{\sffamily-b}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel b.\newline
  \textbf{c} Precision and recall average of 16 tumor and 16 normal single cells sequence at the whole exome level. \newline \footnotesize
  Used thresholds are not comparable between tools.
  They were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
  }
 \label{fig:alt-calling_prec-rec_global}
\end{figure}

The whole genome cell line dataset (Figure~\ref{fig:datasets}a) seems much less challenging than the other dataset: all methods achieve high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec_global}a and \ref{fig:alt-calling_prec-rec}a), at recall rates of 45\% and higher. 
In comparison with all other tools, ProSolo achieves the most striking increases in recall of nearly 10\%.
For example, for a precision above .99, its maximum recall is .766 compared to .705 for MonoVar, .687 for SCIPhI and .610 for SCcaller.
SCAN-SNV achieves up to .992 precision, but only .0001 recall.
This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants\footnote{E.g. see this comment in the issues of the respective repository: \url{https://github.com/parklab/scan-snv/issues/5##issuecomment-628107370}}.

Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset.
This dataset is considerably more challenging than the first one.
See Figures~\ref{fig:alt-calling_prec-rec}b and \ref{fig:alt-calling_prec-rec_global}b for respective results.
For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}b and \ref{fig:alt-calling_prec-rec_global}b).
In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
On this dataset, SCAN-SNV's recall increased to .0016\% at a decreased maximum precision of .897.
Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
At the sites with somatic mutations in single cells, which SCAN-SNV focuses on, this ground truth will instead contain the germline genotype and will incorrectly classify alternative alleles as false positives.
Due to this effect, we also expect the calculated precision of all the other tools to be an underestimate, but against the backdrop of large numbers of sites where the single cells harbour the germline genotype, the effect will be much smaller.
The same ground truth caveat also applies---inversely---for the recall of SCIPhI's sensitive mode and ProSolo's imputation mode.
Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
While both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation, we expect that a number of imputed sites were incorrectly classified as true positives with this germline ground truth.
As this will lead to an overestimation of the recall, we have excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of this dataset---however, their results are nevertheless displayed in all respective figures.

@@

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_per-cell_prec-rec.pdf}
 \caption{
 Cell-specific precision-recall plots across different filtering threshold levels for the whole exome dataset.\newline \footnotesize
  $^\text{\sffamily-b}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline
  Used thresholds are not comparable between tools.
  They were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
 \label{fig:per-cell_prec-rec_Laehnemann2017}
\end{figure}

@@

  \item SCIPhI calls nine heterozygous mutations in 12 or 15 cells (Table~\ref{tab:sciphi-overimputed} and Figure~\ref{fig:somatic-recall}), while the maximum frequency for any of the heterozygous subclonal mutations detected in the deeply sequenced tumor bulk was 0.233, more in the range of the maximum number of six out of 16 cells that \cite{wang_clonal_2014} found any of the subclonal variants in.
  Also, when able to detect the variant, all other tools report a maximum excess of two cells with the mutation compared to \cite{wang_clonal_2014}.
  The only exception is a mutation on chromosome five, where all tools agree on a much larger number of cells with the mutation (9-11) than originally reported by Wang et al. (6), but SCIPhI still clearly exceeds that consensus (15).
  Altogether, these numbers suggest that SCIPhI's model over-imputes mutations by aggregating information across cells.

@@

\subsection{Software and Resources used}
